---
title: "Analysis of in vivo Isotope Labeled Duckweed"
author: "Raven L B Johnson"
date: "2025-02-25"
output:   
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Analysis of in vivo Isotope Labeled Duckweed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

```{r setup, , echo=FALSE, message=FALSE}
library(Cardinal)
setwd("/work/LAS/yjlee-lab/buckm065/Project1_Labeling")
```

## Unlabeled Duckweed Analysis

Normalization and processing is important for consistent MSIi analysis. Here, we import the imzML file and perform TIC normalization and peak binning. We sum to intensities in the pixels to create a TIC image and average the intensities for each m/z value to produce an average mass spectrum for the entire MSI dataset.

```{r}
# import
cntrl <- readMSIData("0A.imzML")

# normalize and bin at 3 mDa
cntrl_normAlign <- cntrl |>
  normalize(method="tic") |>
  peakProcess(tolerance=0.003, units="mz")

# calculate TIC and mean spectrum
cntrl_normAlign <- cntrl_normAlign |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))
```

Below we plot the TIC and mean spectrum.

```{r image-cntrl_tic, fig.height=4, fig.width=6}
image(cntrl_normAlign,"TIC", col=matter::cpal("Cividis"), enhance="hist")
```

```{r image-cntrl_mean, fig.height=5, fig.width=9}
plot(cntrl_normAlign, "Mean", annPeaks=3)
```

We can perform spatial shrunken centroid segmentation on the unlabeled duckweed imaging dataset.

```{r}
cntrl_nobkgrem <- spatialShrunkenCentroids(cntrl_normAlign, 
                                           r = 1, s=2^(1:5), k = 5,
                                           weights = "gaussian")
cntrl_nobkgrem
```

Below, we plot the results of the SSC segmentation without the removal of the tape background.

```{r image-cntrl_seg_bkgd, fig.height=4, fig.width=9}
image(cntrl_nobkgrem, i=1:5, key=TRUE, select=c(5,3,2,1,4), 
      col=c("#B07AA1","#59A14E","#76B7B2","#4F7AA8","#F59AA5"))
```

```{r image-cntrl_regions, fig.height=4, fig.width=6}
cntrl_bkgdrem <- spatialShrunkenCentroids(cntrl_normAlign, 
                                          r = 3, s=16,k=2, weights = "gaussian")
image(cntrl_bkgdrem)
cntrl_normAlign$regions <- makeFactor(
  tissue = (cntrl_bkgdrem@model[["class"]]==2), 
  bkgd = (cntrl_bkgdrem@model[["class"]]==1))
image(cntrl_normAlign, "regions")
```

```{r image-cntrl_bkgd_mean, fig.height=5, fig.width=9}
# subset the background pixels
cntrl_bkgd <- subsetPixels(cntrl_normAlign, regions=="bkgd")

# calculate the TIC and mean spectrum
cntrl_bkgd <- cntrl_bkgd |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))
plot(cntrl_bkgd, "Mean", annPeaks=3)
```

```{r image-cntrl_tissue_mean, fig.height=5, fig.width=9}
# subset the non-background pixels
cntrl_tissue <- subsetPixels(cntrl_normAlign, regions!="bkgd")

# calculate the TIC and mean spectrum for the non-background pixels
cntrl_tissue <-cntrl_tissue |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(cntrl_tissue, "Mean", annPeaks=3)
```

```{r image-cntrl_tissue_ssc, fig.height=4, fig.width=9}
# SSC segmentation of tissue-only region
cntrl_ssc_tissue <- spatialShrunkenCentroids(cntrl_tissue, r=1, s= 2^(1:5),k=4, 
                                            weights = "adaptive")
# plot all models
image(cntrl_ssc_tissue, i=1:5, key=FALSE, select=c(3,4,2,1),
      col=c("#B07AA1","#F28E27","#59A14E","#76B7B2"))
```

```{r image-cntrl_mdl3, fig.height=4, fig.width=9}
# select s = 8 model
cntrl_ssc_tissue_Mdl3 <- cntrl_ssc_tissue[[3]]

# segmentation
image(cntrl_ssc_tissue_Mdl3, key=FALSE, select = c(3,4,2,1),
      col=c("#B07AA1","#F28E27","#59A14E","#76B7B2"))
```

```{r image-cntrl_tstat, fig.height=4, fig.width=9}
# t-statistics
plot(cntrl_ssc_tissue_Mdl3, type="statistic", key=FALSE, 
     scale=TRUE, lwd=2,select = c(3,4,2,1),
     col=c("#B07AA1","#F28E27","#59A14E","#76B7B2"))
```

```{r image-cntrl_cent, fig.height=4, fig.width=9}
# shrunken centroids
plot(cntrl_ssc_tissue_Mdl3, type="centers", key=FALSE, 
     scale=TRUE, lwd=2,select = c(3,4,2,1),
     col=c("#B07AA1","#F28E27","#59A14E","#76B7B2"))
```

```{r image-cntrl_anat, fig.height=4, fig.width=6}
# create a factor for each segment
cntrl_tissue$anatomy <- makeFactor(
  tissueEdge = (cntrl_ssc_tissue_Mdl3@model[["class"]] == 3),
  outerParent = (cntrl_ssc_tissue_Mdl3@model[["class"]] == 4),
  Daughter = (cntrl_ssc_tissue_Mdl3@model[["class"]] == 2),
  buddingPouch = (cntrl_ssc_tissue_Mdl3@model[["class"]] == 1))

# plot the segments
image(cntrl_tissue, "anatomy",
      col=c("#59A14E","#76B7B2","#F28E27","#B07AA1"))
```

```{r image-cntrl_ions_top, fig.height=4, fig.width=9}
# subset the top features from the model
cntrl_top <- topFeatures(cntrl_ssc_tissue_Mdl3)

# plot one ion image for each segment
image(cntrl_tissue, mz=c(483.069,212.843, 465.023,348.868), 
      tolerance=0.002, units="mz", enhance="hist",scale=TRUE,
      col=matter::cpal("Cividis"))
```

```{rimage-cntrl_ions_phosphate, fig.height=4, fig.width=9}
image(cntrl_tissue, mz=c(348.868, 212.852), 
      tolerance=0.002, units="mz", enhance="hist",scale=TRUE,
      col=matter::cpal("Cividis"))
```

## Two-Day 13C-Labeled Duckweed

```{r image-two_mean, fig.height=5, fig.width=9}
twoA <- readMSIData("2dayA_13C_Realigned.imzML")
centroided(twoA) <- FALSE

twoA_normAlign <- twoA |>
  normalize(method="tic") |>
  peakProcess(tolerance=0.002, units="mz")
twoA_normAlign <- twoA_normAlign |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(twoA_normAlign, "Mean", annPeaks=3)
```

```{r image-two_regions, fig.height=4, fig.width=6}
twoA_bkgdrem <- spatialShrunkenCentroids(twoA_normAlign, r = 3, s=0,k=3, 
                                         weights = "gaussian")
twoA_normAlign$regions <- makeFactor(
  tissue = (twoA_bkgdrem@model[["class"]]!=3),
  bkgd = (twoA_bkgdrem@model[["class"]]==3))

image(twoA_normAlign, "regions")
```

```{r image-two_bkgd_mean, fig.height=5, fig.width=9}
twoA_bkgd <- subsetPixels(twoA_normAlign, regions=="bkgd")
twoA_bkgd <- twoA_bkgd |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(twoA_bkgd, "Mean", annPeaks=3)
```

```{r image-two_tissue_mean, fig.height=4, fig.width=9}
twoA_tissue <- subsetPixels(twoA_normAlign, regions!="bkgd")
twoA_tissue <- twoA_tissue |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(twoA_tissue, "Mean", annPeaks=2)
```

```{r image-two_ssc, fig.height=4, fig.width=9}
twoA_ssc_tissue <- spatialShrunkenCentroids(twoA_tissue, r = 1, s= 2^(0:5),k=4, 
                                            weights = "gaussian")
twoA_ssc_tissue
image(twoA_ssc_tissue, i=c(1:6), col=c("#F28E27","#E05759","#59A14E","#76B7B2"))
```

```{r image-two_mdl4, fig.height=4, fig.width=6}
twoA_ssc_tissue_Mdl4 <- twoA_ssc_tissue[[4]]
image(twoA_ssc_tissue_Mdl4, key=FALSE,
      col = c("#F28E27","#E05759","#59A14E","#76B7B2"))
```

```{r image-two_tstat, fig.height=4, fig.width=9}
plot(twoA_ssc_tissue_Mdl4, type="statistic", key=FALSE, scale=TRUE, 
     lwd=2, col = c("#F28E27","#E05759","#59A14E","#76B7B2"))
```

```{r image-two_cent, fig.height=4, fig.width=9}
plot(twoA_ssc_tissue_Mdl4, type="centers", key=FALSE, scale=TRUE, 
     lwd=2, col = c("#F28E27","#E05759","#59A14E","#76B7B2"))
```

```{r image-two_anat, fig.height=4, fig.width=6}
twoA_top <- topFeatures(twoA_ssc_tissue_Mdl4)
twoA_tissue$anatomy <- makeFactor(
  tissueEdge = (twoA_ssc_tissue_Mdl4@model[["class"]] == 3),
  Daughter = (twoA_ssc_tissue_Mdl4@model[["class"]] == 1),
  innerParent = (twoA_ssc_tissue_Mdl4@model[["class"]] == 2),
  outerParent = twoA_ssc_tissue_Mdl4@model[["class"]] == 4)
image(twoA_tissue, "anatomy",
      col=c("#59A14E","#F28E27","#E05759","#76B7B2"))
```

## Three-Day 13C-Labeled Duckweed

```{r iamge-three_mean, fig.height=4, fig.width=9}
threeC <-readMSIData("3dayC_13C_Realigned.imzML")
centroided(threeC) <- FALSE

threeC_normAlign <- threeC |>
  normalize(method="tic") |> #TIC Normalization
  peakProcess(tolerance=0.002, units="mz")

threeC_normAlign<- threeC_normAlign |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(threeC_normAlign, "Mean", annPeaks=3)
```

```{r image-three_TIC, fig.height=4, fig.width=6}
image(threeC_normAlign, "TIC", enhance="hist",col=matter::cpal("Cividis"))
```

```{r image-three_regions, fig.height=4, fig.width=6}
threeC_bkgdrem <- spatialShrunkenCentroids(threeC_normAlign, r = 2, s= 8,k=3, 
                                           weights = "gaussian")

threeC_normAlign$regions <- makeFactor(
  tissue = (threeC_bkgdrem@model[["class"]]!=1),
  bkgd = (threeC_bkgdrem@model[["class"]]==1))

image(threeC_normAlign, "regions")
```

```{r image-three_bkgd_mean, fig.height=5, fig.width=9}
threeC_bkgd <- subsetPixels(threeC_normAlign, regions == "bkgd")

threeC_bkgd <- threeC_bkgd |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(threeC_bkgd, "Mean", annPeaks=4)
```

```{r image-three_tissue_mean, fig.height=5, fig.width=9}
threeC_tissue <- subsetPixels(threeC_normAlign, regions == "tissue")

threeC_tissue <- threeC_tissue |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(threeC_tissue, "Mean", annPeaks=2)
```

```{r image- three_ssc, fig.height=4, fig.width=9}
threeC_ssc_tissue <- spatialShrunkenCentroids(threeC_tissue, 
                                              r=2, k = 5, s= 2^(0:5), 
                                              weights="gaussian")
image(threeC_ssc_tissue, i = 1:6, key=FALSE, 
      col = c("#E05759","#76B7B2","#F28E27","#EDC947","#B07AA1"))
```

```{r image-three_tstat, fig.height=4, fig.width=9}
threeC_ssc_tissue_Mdl4 <- threeC_ssc_tissue[[4]]
plot(threeC_ssc_tissue_Mdl4, type = "statistic", lwd = 2, key=FALSE, 
     col = c("#E05759","#76B7B2","#F28E27","#EDC947","#B07AA1"))
```

```{r image-three_cent, fig.height=4, fig.width=9}
plot(threeC_ssc_tissue_Mdl4, type = "centers", lwd = 2, scale = TRUE, key=FALSE, 
     col = c("#E05759","#76B7B2","#F28E27","#EDC947","#B07AA1"))
```

```{r image-three_anat, fig.height=4, fig.width=6}
threeC_top <- topFeatures(threeC_ssc_tissue_Mdl4)

threeC_tissue$anatomy <- makeFactor(
  innerDaughter = (threeC_ssc_tissue_Mdl4@model[["class"]] == 3),
  outerDaughter = (threeC_ssc_tissue_Mdl4@model[["class"]] == 4),
  innerParent = (threeC_ssc_tissue_Mdl4@model[["class"]] == 1),
  buddingPouch = (threeC_ssc_tissue_Mdl4@model[["class"]] == 5),
  outerParent = threeC_ssc_tissue_Mdl4@model[["class"]] == 2)
image(threeC_tissue, "anatomy",
      col=c("#F28E27","#EDC947","#E05759","#B07AA1","#76B7B2"))

```

## Five-Day D-Labeled Duckweed

```{r image- five_mean, fig.height=5, fig.width=9}
fiveB <-readMSIData("5dayB_2H.imzML")

fiveB_normAlign<- fiveB|>
  normalize(method="tic") |> #TIC Normalization
  peakProcess(tolerance=0.002, units="mz", filterFreq = 0)

fiveB_normAlign <- fiveB_normAlign |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))
plot(fiveB_normAlign, "Mean", annPeaks = 2)
```

```{r image-five_TIC, fig.height=4, fig.width=6}
image(fiveB_normAlign, "TIC", enhance="hist", col=matter::cpal("Cividis"))
```

### Lipid Analysis

```{r image-five_lip_mean, fig.height=5, fig.width=9}
fiveB_lipids <- subsetFeatures(fiveB_normAlign,800 < mz, mz < 1200)
fiveB_lipids <- normalize(fiveB_lipids, method="tic")|>
  process()

fiveB_lipids <- fiveB_lipids |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))

plot(fiveB_lipids, "Mean", annPeaks = 2)
```

```{r image-five_lip_regions, fig.height=4, fig.width=6}
fiveB_lipids_ssc <- spatialShrunkenCentroids(fiveB_lipids, r = 2, 
                                                s = 0, k = 2)
fiveB_lipids_ssc
fiveB_lipids$regions <- makeFactor(tissue = (fiveB_lipids_ssc@model[["class"]]!=1),
                                  bkgd = (fiveB_lipids_ssc@model[["class"]]==1))
image(fiveB_lipids,"regions")
```

```{r image-five_bkgd_mean, fig.height=5, fig.width=9}
fiveB_bkgd_lipids <- subsetPixels(fiveB_lipids, regions == "bkgd")
fiveB_bkgd_lipids <- fiveB_bkgd_lipids |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))
plot(fiveB_bkgd_lipids, "Mean", annPeaks = 2)
```

```{r image-five_lip_tstat, fig.height=4, fig.width=6}
fiveB_tissue_lipids <- subsetPixels(fiveB_lipids, regions == "tissue")
fiveB_tissue_lipids_ssc <- spatialShrunkenCentroids(fiveB_tissue_lipids, r=3, 
                                            k = 2, s=8, 
                                            weights="gaussian")
fiveB_lipid_top <- topFeatures(fiveB_tissue_lipids_ssc)
plot(fiveB_tissue_lipids_ssc, type="statistic",lwd = 2,
     col=c("#76B7B2","#F28E27"), key = FALSE)
```

```{r image-five_lip_cent, fig.height=4, fig.width=6}
plot(fiveB_tissue_lipids_ssc, type="centers",lwd = 2,
     col=c("#76B7B2","#F28E27"), key =FALSE)
```

```{r image-five_lip_anat, fig.height=4, fig.width=6}
fiveB_tissue_lipids$anatomy <- makeFactor(
  Parent = (fiveB_tissue_lipids_ssc@model[["class"]] == 1),
  Daughter = (fiveB_tissue_lipids_ssc@model[["class"]] == 2))
image(fiveB_tissue_lipids, "anatomy", col=c("#76B7B2","#F28E27"))
```

### Whole Mass Range Analysis

```{r image-five_bkgd_mean, fig.height=5, fig.width=9}
fiveB_normAlign$regions <- makeFactor(tissue = (fiveB_tissue_lipids_ssc@model[["class"]]!=1),
                                  bkgd = (fiveB_tissue_lipids_ssc@model[["class"]]==1))

fiveB_tissue <- subsetPixels(fiveB_normAlign, regions == "tissue")
fiveB_bkgd <- subsetPixels(fiveB_normAlign, regions == "bkgd")

fiveB_bkgd <- fiveB_bkgd |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))
plot(fiveB_bkgd, "Mean", annPeaks = 2)
```

```{r image-five_tissue_mean, fig.height=4, fig.width=9}
fiveB_tissue <- fiveB_tissue |>
  summarizePixels(c(TIC="sum")) |>
  summarizeFeatures(c(Mean = "mean"))
plot(fiveB_tissue, "Mean", annPeaks = 1)
```

```{r image-five_tissue_ssc, fig.height=4, fig.width=9}
fiveB_tissue_ssc <- spatialShrunkenCentroids(fiveB_tissue, r=3, k=5, s=2^(0:5), 
                                             weights="adaptive")
image(fiveB_tissue_ssc, i = 1:6, 
      col=c("#59A14E","#76B7B2","#B07AA1","#E05759","#F28E27"))
```

```{r image-five_tstat, fig.height=4, fig.width=9}
fiveB_tissue_Mdl4 <- fiveB_tissue_ssc[[4]]
fiveB_tissue_top <- topFeatures(fiveB_tissue_Mdl4)

plot(fiveB_tissue_Mdl4, type="statistic", lwd = 2, key = FALSE, 
     col=c("#59A14E","#76B7B2","#B07AA1","#E05759","#F28E27") )
```

```{r image-five_cent, fig.height=4, fig.width=9}
plot(fiveB_tissue_Mdl4, type="centers", lwd = 2, key = FALSE, 
     col=c("#59A14E","#76B7B2","#B07AA1","#E05759","#F28E27") )
```

```{r image-five_anat, fig.height=4, fig.width=6}
fiveB_tissue$anatomy <- makeFactor(
  tissueEdge = (fiveB_tissue_Mdl4@model[["class"]] == 1),
  outerParent = (fiveB_tissue_Mdl4@model[["class"]] == 2),
  buddingPouch = (fiveB_tissue_Mdl4@model[["class"]] == 3),
  innerParent = (fiveB_tissue_Mdl4@model[["class"]] == 4),
  Daughter = (fiveB_tissue_Mdl4@model[["class"]] == 5))
image(fiveB_tissue, "anatomy",
      col=c("#59A14E","#76B7B2","#B07AA1","#E05759","#F28E27"))
```

## Session information

```{r}
sessionInfo()
```
